\begin{exercise}{0}
  Which of the following statements hold?
  \begin{enumerate}
    \item Deterministic finite automata (DFA) are strictly less expressive than regular expressions.
    \item Non-deterministic finite automata (NFA) are strictly more expressive than DFA.
    \item The regular languages are closed under:
      \begin{enumerate}
    	 \item union,
    	 \item intersection,
    	 \item complement,
    	 \item concatenation,
    	 \item Kleene closure.
       \end{enumerate}
    \item Context Free Languages (CFL) are closed under:
      \begin{enumerate}
        \item union,
    	\item intersection,
    	\item complement,
    	\item concatenation,
    	\item Kleene closure.
      \end{enumerate}
    \item DCFL is the set of context free languages that are accepted by deterministic push down automata. Is DCFL = CFL?
  \end{enumerate}
\end{exercise}


\begin{solution}
  \begin{enumerate}
    \item No. One can be effectively converted to another.
    \item No. NFA can be converted to a DFA using the powerset construction (Rabin-Scott construction).
    \item Yes for all. Each operation can be implemented by manipulating a DFA or NFA.
    \item Context Free Languages (CFL) are closed under the following:
      \begin{enumerate}
	    \item Yes. $S_1$ and $S_2$ are starting Non-terminal for two Grammar. $S \to S_1 \mid S_2.$
	    \item No. $L_1 = \{a^ib^ic^j \mid i,j\in\mathbb{N}\}$ and $L_2 = \{a^ib^jc^j \mid i,j\in \mathbb{N}\}$
	    \item No. CFLs are closed under union and if they were closed under complement, they would also be closed under intersection.
	    \item Yes.  $S \to S_1  S_2.$
	    \item Yes. Let $S$ be the starting non-terminal. New Grammar $S' \to S S' \mid \varepsilon$.
	\end{enumerate}
    \item No. Just an intuitive example: Palindromes are in CFL but not in DCFL.
  \end{enumerate}
\end{solution}
