\documentclass[a4paper]{article}

\input{../commands}

\begin{document}

\header{January 20th}

%\section*{General Remarks}
%\begin{itemize}
%  \item This programming exercise is optional and provides bonus points for the exercise class.
%\end{itemize}

\begin{exercise}{10}
  We now finish the implementation of our \icompiler{} by generating the corresponding \texttt{Jasmin} code for our parsed program.

  As stated on the \href{http://jasmin.sourceforge.net/}{Jasmin Webpage}:\\
  ``\emph{Jasmin is an assembler for the Java Virtual Machine. It takes ASCII descriptions of Java classes, written in a simple assembler-like syntax using the Java Virtual Machine instruction set. It converts them into binary Java class files, suitable for loading by a Java runtime system.}''

  Our compiler generates code for the Jasmin language from the parsed \textit{While} language. As a recap the grammar for the \textit{While} language is given as follows:
  \begin{center}
  \begin{tabular}{lll}
    start       &$\to$ & program EOF\\
    program     &$\to$ & statement program $\mid$ statement\\
    statement   &$\to$ & declaration SEM $\mid$ assignment SEM $\mid$ branch $\mid$ loop $\mid$ out SEM\\
    declaration &$\to$ & INT ID\\
    assignment  &$\to$ & ID ASSIGN expr $\mid$ ID ASSIGN READ LBRAC RBRAC\\
    out         &$\to$ & WRITE LBRAC expr RBRAC $\mid$ WRITE LBRAC STRING RBRAC\\
    branch      &$\to$ & IF LBRAC guard RBRAC LCBRAC program RCBRAC $\mid$\\
                &      & IF LBRAC guard RBRAC LCBRAC program RCBRAC\\
                &      & ELSE LCBRAC program RCBRAC\\
    loop        &$\to$ & WHILE LBRAC guard RBRAC LCBRAC program RCBRAC\\
    expr        &$\to$ & NUMBER $\mid$ ID $\mid$ subexpr\\
    subexpr     &$\to$ & expr PLUS expr $\mid$ expr MINUS expr $\mid$ expr TIMES expr $\mid$ expr DIV expr\\
                &      & $\mid$ expr MOD expr\\
    guard       &$\to$ & relation $\mid$ subguard $\mid$ NOT LBRAC guard RBRAC\\
    subguard    &$\to$ & guard AND guard $\mid$ guard OR guard\\
    relation    &$\to$ & expr LT expr $\mid$ expr LEQ expr $\mid$ expr EQ expr $\mid$ expr NEQ expr $\mid$ expr GEQ expr\\
                &      & $\mid$ expr GT expr
  \end{tabular}
  \end{center}

Implement \texttt{generator.JasminGenerator.translateProgram(ASTNode)} which given the \emph{program} node in an abstract syntax tree returns the generated Jasmin Code as a string.

\emph{Hint: It is a good approach to write methods for every language construct and call them recursively (similar to a recursive descent parser). Once you get the idea, it is actually less effort than you might think! And do not forget to zero-initialize newly declared variables!}

Methods for generating \texttt{Jasmin} code for the main class, for writing to and reading from the console are already provided. You should also implement the method \texttt{translateExpr(ASTNode)} which translates an \emph{expression} into \texttt{Jasmin} code and is used in the method for writing to the console.

Here are some helpful resources for the \texttt{Jasmin} language:
  \begin{itemize}
    \item Jasmin main page: \href{http://jasmin.sourceforge.net/}{http://jasmin.sourceforge.net/}
    \item Jasmin user guide: \href{http://jasmin.sourceforge.net/guide.html}{http://jasmin.sourceforge.net/guide.html}
    \item List of machine instructions for Jasmin: \href{http://jasmin.sourceforge.net/instructions.html}{http://jasmin.sourceforge.net/instructions.html}
    \item Explanation of the machine instructions: \href{https://docs.oracle.com/javase/specs/jvms/se8/html/jvms-6.html}{https://docs.oracle.com/javase/specs/jvms/se8/html/jvms-6.html}
  \end{itemize}

  To test your implementation you can write code in the \textit{WHILE} language and run it through our compiler with:
  \begin{lstlisting}[basicstyle=\ttfamily,numbers=none]
$java -cp bin Main tests/gcd.txt --out gcd.j
   \end{lstlisting}
  The generated code is written to the given filename (in this example \texttt{gcd.j}). Next you can use \texttt{Jasmin} to build an executable Java class file:
  \begin{lstlisting}[basicstyle=\ttfamily,numbers=none]
$java -jar lib/jasmin.jar gcd.j
   \end{lstlisting}
You can execute the generated class file as a Java program and observe its behavior:
  \begin{lstlisting}[basicstyle=\ttfamily,numbers=none]
$java gcd
42
27
GCD: 
3
   \end{lstlisting}

  After finishing this exercise you now have your own simple compiler for Java code!
\end{exercise}

\begin{solution}
  For the solution see the code in \texttt{i2CompilerSolution}.
\end{solution}

\end{document}
